Perturbing SU($n$)-symmetric interactions

### 2020-03-09 collective_sun/sun_perturbation.tex
minor change in notation

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the
permutationally symmetric manifold $\M_0$ by the two-body perturbation
$\V_2$ in \eqref{eq:perturbations} is simply the restriction of $\V_2$
onto $\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQPS \sum_O \sum_{p<q} w_{Opq} O_{1,2}
  \EQPS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-spin
operator $O_{pq}$.  If the two-spin operator
$O=X\otimes Y+Y\otimes X$, then
\begin{align}
  H_2^{(1)} \EQPS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} + \col{Y}\,\col{X} - \col{XY} - \col{YX}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary permutationally symmetric
state $\ket\psi\in\M_0$ and expand (see Section \ref{sec:eigenstates})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \sum_O \sum_{p<q}
  \sp{\sum_k \p{h_{pk} w_{Okq} + h_{qk} w_{Opk}}
  - \bar h_{pq} w_{Opq}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_p + h_q - 2 h_{pq},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ (indexed by $\Delta$) is an eigenvector
of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h}_2 \c \v{\m w} = \Delta \v{\m w},
  \label{eq:V2_eig}
\end{align}
where
\begin{align}
  \check{\m h}_2
  \equiv \sum_{p<q} \ket{\set{p,q}} \sp{\sum_{k\notin\set{p,q}}
    \p{h_{pk} \bra{\set{k,q}} + h_{qk} \bra{\set{p,k}}}
    - \bar h_{pq} \bra{\set{p,q}}}.
  \label{eq:V2_eig_mat}
\end{align}
We explicitly exclude $k\in\set{p,q}$ in \eqref{eq:V2_eig_mat} because
these terms vanish in \eqref{eq:H_V2_psi} from the fact that
$h_{kk}=w_{Okk}=0$.  Solving the general eigenvalue problem in
\eqref{eq:V2_eig} requires diagonalizing the
${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$ matrix
$\check{\m h}_2$.  When both he interaction Hamiltonian $H_0$
translationally invariant, however, this eigenvalue problem can be
reduced to that of diagonalizing an $\sim N\times N$ matrix; we
perform this reduction in Section \ref{sec:eigenstates_TI}.

@@

  = E_0 - \bar h_k,

@@

  \bar h_k \equiv \sum_{p\in k}\p{h_p - \sum_{q\in k}h_{pq}},

@@

  \sum_{k\in\C_N\p{M}} \p{E_0 - \bar h_k} w_k O_k \ket\psi

@@

  - \sum_{k\in\C_N\p{M}} \p{\bar{\m h} * \m w}_k O_k \ket\psi,

@@

where $\bar{\m h}$ is a tensor with entries $\bar h_k$; and

@@

  \p{\m h\circ\m w - \bar{\m h} * \m w}_k O_k \ket\psi.

@@

  \m h\circ\m w - \bar{\m h} * \m w = \Delta \m w.

@@

$\bar h_k=h_k$, so

@@

    - \bar h_{pq} w_{pq}}

@@

where $\bar h_{pq}=h_p+h_q-2h_{pq}$ with $h_p\equiv\sum_qh_{pq}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Translational invariance}
\label{sec:eigenstates_TI}

@@

Our task is essentially to express the multi-body eigenvalue problem
in \eqref{eq:H_VM_psi_eig} in a basis that obeys both permutational
symmetry and translational invariance.  To this end, we first expand
the coupling tensor $\m w$ in a complete basis for symmetric tensors:
\begin{align}
  \m w = \sum_{k\in\C_N\p{M}} w_k \m m_k,
  &&
  \m m_k \equiv \sum_{\sigma\in P_M} \m e_{\sigma\p{k}},
  \label{eq:sym_tensor}
\end{align}
where $P_M$ is the set of all permutations of $M$ symbols, and
$\m e_\ell$ is a unit tensor with one in the entry indexed by $\ell$
and zeros everywhere else.  On a $D$-dimensional lattice of linear
size $L$, translational invariance implies that the coefficients $w_k$
remain the same if all spins indexed by $k$ are shifted by the same
displacement $d\in\ZZ_L^D$.  To account for this invariance, we denote
a translation of all spins in $k$ by a displacement $d\in\ZZ_L^D$ by
$k_d$, and define the translation-symmetrized basis tensors
\begin{align}
  \m m_k^{\t{TS}}
  \equiv \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \m m_{k_d}
  = \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \sum_{\sigma\in P_M}
  \m e_{\sigma(k_d)},
\end{align}
where the normalization factor $\N_k^{\t{TS}}$ enforces that the
entries of $\m m_k^{\t{TS}}$ are all either zero or one, which
otherwise might not occur if some translations of $k$ results in the
same choice of spins.  We then define the set $\T_{LD}\p{M}$ of
equivalence classes in $\C_N\p{M}$ under arbitrary translations on
$D$-dimensional lattice of linear size $L$, and expand
\begin{align}
  \m w = \sum_{k\in\T_{LD}\p{M}} w_k \m m_k^{\t{TS}}.
  \label{eq:sym_shift_tensor}
\end{align}
Having expanded $\m w$ in a basis that respects permutational symmetry
and translational invariance, we need to express the eigenvalue
equation \eqref{eq:H_VM_psi_eig} in the same basis.  In order to do
so, we label spins $p,q$ by the position of the corresponding lattice
site in $\ZZ_L^D$, and use translational invariance of the couplings
$\m h$ to expand
\begin{align}
  \m h \equiv \sum_{p,q\in\ZZ_L^D} h_{pq} \op{p}{q}
  = \sum_{q,d\in\ZZ_L^D} h_{q+d,q} \op{q+d}{q}
  = \sum_{d\in\ZZ_L^D} h_{d,0} \m s_d,
  &&
  \m s_d \equiv \sum_{q\in\ZZ_L^D} \op{q+d}{d}.
\end{align}
Starting from this form of $\m h$ and the expansion of $\m w$ in
\eqref{eq:sym_tensor}, we can work out that
\begin{align}
  \m h\circ\m w
  = \sum_{k\in\T_{LD}\p{M}} \sum_{d\in\ZZ_L^D} \sum_{a=1}^M w_{k_{d;a}}
  h_{d,0} w_{k_{d;a}} \m m_k^{\t{TS}},
\end{align}
where $k_{d;a}$ denotes the spins acquired by displacing the $a$-th
spin of $k$ by $d$.  With a bit of additional work, we can also
simplify
\begin{align}
  \bar{\m h}
  = \sum_{k\in\T_{LD}\p{M}} \sp{M h_0 - \sum_{p,q\in k} h_{pq}}
  \m m_k^{\t{TS}},
  &&
  h_0 \equiv \sum_{d\in\ZZ_L^D} h_{d,0},
\end{align}
which implies that
\begin{align}
  \bar{\m h} * \m w = M h_0 \m w
  - \sum_{k\in\T_{LD}\p{M}} \sum_{p,q\in k} h_{pq} w_k \m m_k^{\t{TS}}.
\end{align}
Putting everything together, the generalized eigenvalue problem in
\eqref{eq:H_VM_psi_eig} can be written in the form
\begin{align}
  \sum_{k\in\T_{LD}\p{M}} \sp{\sum_{p,q\in k} h_{pq} w_k
    + \sum_{d\in\ZZ_L^D} h_{d,0} \sum_{a=1}^M w_{k_{d;a}}}
  \m m_k^{\t{TS}}
  = \p{M h_0 + \Delta} \m w,
\end{align}
or equivalently
\begin{align}
  \check{\m h} \c \v{\m w} = \p{M h_0 + \Delta} \v{\m w},
  \label{eq:H_VM_eig_simp}
\end{align}
where
\begin{align}
  \check{\m h} \equiv
  \sum_{k\in\T_{LD}\p{M}} \ket{k}
  \sp{\sum_{p,q\in k} h_{pq} \bra{k}
    + \sum_{d\in\ZZ_L^D} h_{d,0} \sum_{a=1}^M \bra{k_{d;a}}},
  &&
  \v{\m w} \equiv \sum_{k\in\T_{LD}\p{M}} w_k \ket{k},
\end{align}
and, to keep notation simple, we define $\ket{k}=0$ for all $k$ with
repeated values, and enforce $\ket{j}=\ket{k}$ if spins $j$ and spins
$k$ are the same up to a global translation.  The matrix
$\check{\m h}$ has dimensions
$\abs{\T_{LD}\p{M}}\times\abs{\T_{LD}\p{M}}$, where
$\abs{\T_{LD}\p{M}} \approx { N-1 \choose M-1 }/M \sim N^{M-1}$.

>>>>

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the
permutationally symmetric manifold $\M_0$ by the two-body perturbation
$\V_2$ in \eqref{eq:perturbations} is simply the restriction of $\V_2$
onto $\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQPS \sum_O \sum_{p<q} w_{Opq} O_{1,2}
  \EQPS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-spin
operator $O_{pq}$.  If the two-spin operator
$O=X\otimes Y+Y\otimes X$, then
\begin{align}
  H_2^{(1)} \EQPS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} + \col{Y}\,\col{X} - \col{XY} - \col{YX}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary permutationally symmetric
state $\ket\psi\in\M_0$ and expand (see Section \ref{sec:eigenstates})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \sum_O \sum_{p<q}
  \sp{\sum_k \p{h_{pk} w_{Okq} + h_{qk} w_{Opk}}
    - \tilde h_{pq} w_{Opq}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \tilde h_{pq} \equiv h_p + h_q - 2 h_{pq},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ (indexed by $\Delta$) is an eigenvector
of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h}_2 \c \v{\m w} = \Delta \v{\m w},
  \label{eq:V2_eig}
\end{align}
where
\begin{align}
  \check{\m h}_2
  \equiv \sum_{p<q} \ket{\set{p,q}} \sp{\sum_{k\notin\set{p,q}}
    \p{h_{pk} \bra{\set{k,q}} + h_{qk} \bra{\set{p,k}}}
    - \tilde h_{pq} \bra{\set{p,q}}}.
  \label{eq:V2_eig_mat}
\end{align}
We explicitly exclude $k\in\set{p,q}$ in \eqref{eq:V2_eig_mat} because
these terms vanish in \eqref{eq:H_V2_psi} from the fact that
$h_{kk}=w_{Okk}=0$.  Solving the general eigenvalue problem in
\eqref{eq:V2_eig} requires diagonalizing the
${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$ matrix
$\check{\m h}_2$.  When both he interaction Hamiltonian $H_0$
translationally invariant, however, this eigenvalue problem can be
reduced to that of diagonalizing a matrix of dimensions
$\sim N\times N$; we perform this reduction in Section
\ref{sec:exploiting_symmetries}.

@@

  = E_0 - \tilde h_k,

@@

  \tilde h_k \equiv \sum_{p\in k}\p{h_p - \sum_{q\in k}h_{pq}},

@@

  \sum_{k\in\C_N\p{M}} \p{E_0 - \tilde h_k} w_k O_k \ket\psi

@@

  - \sum_{k\in\C_N\p{M}} \p{\tilde{\m h} * \m w}_k O_k \ket\psi,

@@

where $\tilde{\m h}$ is a tensor with entries $\tilde h_k$; and

@@

  \p{\m h\circ\m w - \tilde{\m h} * \m w}_k O_k \ket\psi.

@@

  \m h\circ\m w - \tilde{\m h} * \m w = \Delta \m w.

@@

$\tilde h_k=h_k$, so

@@

    - \tilde h_{pq} w_{pq}}

@@

where $\tilde h_{pq}=h_p+h_q-2h_{pq}$ with $h_p\equiv\sum_qh_{pq}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Translational invariance}

@@

Our task is essentially to express the multi-body eigenvalue problem
in \eqref{eq:H_VM_psi_eig} in a basis that obeys both permutational
symmetry and translational invariance.  To this end, we first expand
the coupling tensor $\m w$ in a complete basis for symmetric tensors:
\begin{align}
  \m w = \sum_{k\in\C_N\p{M}} w_k \m m_k,
  &&
  \m m_k \equiv \sum_{\sigma\in P_M} \m e_{\sigma\p{k}},
  \label{eq:sym_tensor}
\end{align}
where $P_M$ is the set of all permutations of $M$ symbols, and
$\m e_\ell$ is a unit tensor with one in the entry indexed by $\ell$
and zeros everywhere else.  On a $D$-dimensional lattice of linear
size $L$, translational invariance implies that the coefficients $w_k$
remain the same if all spins indexed by $k$ are shifted by the same
displacement $d\in\ZZ_L^D$.  To account for this invariance, we denote
a translation of all spins in $k$ by a displacement $d\in\ZZ_L^D$ by
$k_d$, and define the translation-symmetrized basis tensors
\begin{align}
  \m m_k^{\t{TS}}
  \equiv \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \m m_{k_d}
  = \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \sum_{\sigma\in P_M}
  \m e_{\sigma(k_d)},
\end{align}
where the normalization factor $\N_k^{\t{TS}}$ enforces that the
entries of $\m m_k^{\t{TS}}$ are all either zero or one, which
otherwise might not occur if some translations of $k$ results in the
same choice of spins.  We then define the set $\T_{LD}\p{M}$ of
equivalence classes in $\C_N\p{M}$ under arbitrary translations on
$D$-dimensional lattice of linear size $L$, and expand
\begin{align}
  \m w = \sum_{k\in\T_{LD}\p{M}} w_k \m m_k^{\t{TS}}.
  \label{eq:sym_shift_tensor}
\end{align}
Having expanded $\m w$ in a basis that respects permutational symmetry
and translational invariance, we need to express the eigenvalue
equation \eqref{eq:H_VM_psi_eig} in the same basis.  In order to do
so, we label spins $p,q$ by the position of the corresponding lattice
site in $\ZZ_L^D$, and use translational invariance of the couplings
$\m h$ to expand
\begin{align}
  \m h \equiv \sum_{p,q\in\ZZ_L^D} h_{pq} \op{p}{q}
  = \sum_{q,d\in\ZZ_L^D} h_{q+d,q} \op{q+d}{q}
  = \sum_{d\in\ZZ_L^D} h_{d,0} \m s_d,
  &&
  \m s_d \equiv \sum_{q\in\ZZ_L^D} \op{q+d}{d}.
\end{align}
Starting from this form of $\m h$ and the expansion of $\m w$ in
\eqref{eq:sym_tensor}, we can work out that
\begin{align}
  \m h\circ\m w
  = \sum_{k\in\T_{LD}\p{M}} \sum_{d\in\ZZ_L^D} \sum_{a=1}^M w_{k_{d;a}}
  h_{d,0} w_{k_{d;a}} \m m_k^{\t{TS}},
\end{align}
where $k_{d;a}$ denotes the spins acquired by displacing the $a$-th
spin of $k$ by $d$.  With a bit of additional work, we can also
simplify
\begin{align}
  \tilde{\m h}
  = \sum_{k\in\T_{LD}\p{M}} \sp{M h_0 - \sum_{p,q\in k} h_{pq}}
  \m m_k^{\t{TS}},
  &&
  h_0 \equiv \sum_{d\in\ZZ_L^D} h_{d,0},
\end{align}
which implies that
\begin{align}
  \tilde{\m h} * \m w = M h_0 \m w
  - \sum_{k\in\T_{LD}\p{M}} \sum_{p,q\in k} h_{pq} w_k \m m_k^{\t{TS}}.
\end{align}
Putting everything together, the generalized eigenvalue problem in
\eqref{eq:H_VM_psi_eig} can be written in the form
\begin{align}
  \sum_{k\in\T_{LD}\p{M}} \sp{\sum_{p,q\in k} h_{pq} w_k
    + \sum_{d\in\ZZ_L^D} h_{d,0} \sum_{a=1}^M w_{k_{d;a}}}
  \m m_k^{\t{TS}}
  = \p{M h_0 + \Delta} \m w,
\end{align}
or equivalently
\begin{align}
  \check{\m h} \c \v{\m w} = \p{M h_0 + \Delta} \v{\m w},
  \label{eq:H_VM_eig_simp}
\end{align}
where
\begin{align}
  \check{\m h} \equiv
  \sum_{k\in\T_{LD}\p{M}} \ket{k}
  \sp{\sum_{p,q\in k} h_{pq} \bra{k}
    + \sum_{d\in\ZZ_L^D} h_{d,0} \sum_{a=1}^M \bra{k_{d;a}}},
  &&
  \v{\m w} \equiv \sum_{k\in\T_{LD}\p{M}} w_k \ket{k},
\end{align}
and, to keep notation simple, we define $\ket{k}=0$ for all $k$ with
repeated values, and enforce $\ket{j}=\ket{k}$ if spins $j$ and spins
$k$ are the same up to a global translation.  The matrix
$\check{\m h}$ has dimensions
$\abs{\T_{LD}\p{M}}\times\abs{\T_{LD}\p{M}}$, where
$\abs{\T_{LD}\p{M}} \approx { N-1 \choose M-1 }/M \sim N^{M-1}$.
